Normalization by Evaluation for Sized Dependent Types

### 2017-06-10 nbe/icfp17.tex
[ ICFP 17 ] DOIs for all references (where available)

Sized types \cite{hughesParetoSabry:popl96} delegate the checking for structural descent to the type system by annotating data types with a size parameter.  The type checker can then ensure that in recursive calls the size goes down, certifying termination.  In the simplest setting \cite{gimenez:typeBased,abel:lmcs07}, the size is just an upper bound on the tree height of the data structure; however, more sophisticated size annotations have also been considered \cite{xi:terminationHOSC,blanqui:rta04}.  Most sized type systems are non-dependent \cite{amadio:guardcondition,abelPientka:jfp16,bartheGregoireRiba:lernet08,bartheGregoireRiba:csl08,blanquiRiba:lpar06,dalLagoGrellois:probabilisticSizedTypes2017}, yet the combination of sized and dependent types has been studied as well \cite{blanqui:csl05,bartheGregoirePastawski:lpar06,gregoireSacchini:lpar10,sacchini:lics13,sacchini:flops14}.  However, to the best of our knowledge, no study combines \emph{higher-rank size polymorphism} with full-fledged dependent types.\footnote{\citet{xi:terminationHOSC} has first-class size polymorphism, but only indexed types, no universes or large eliminations.}

@@

The central technical observation is that reflection and reification do not need the precise type $T$, they work the same for any \emph{shape} $S$ of $T$.
We managed, while not introducing a new syntax for shapes, to define a relation $T \shape S$ on type values stating that type $S$ qualifies as shape for type $T$.
Shapes unfold dynamically during reflection and reification.
For example, when reflecting a variable $x$ into the polymorphic function type $\forall i.\, F\,i$ where $F\,i = \Nat\,i \to \Nat\,i$, we obtain
$(\up{\forall i.\/ F\,i}x)\,i = \up{F\,i} (x\,i)$ for size $i$ and
$(\up{\forall i.\/ F\,i}x)\,j = \up{F\,j} (x\,j)$ for size $j$.  The new types $F\,i$ and $F\,j$ we reflect at are no longer equal (and they are not subtypes of each other), but they still have the same shape, $\Nat\_ \to \Nat\_$.  This means they will still move in lock-step in respect to $\eta$-expansion, which is sufficient to prove NbE correct for judgmental equality.
We call the enabling property of $F$ \emph{shape irrelevance}, meaning that for any pair $i$, $j$ of legal arguments, $F\,i$ and $F\,j$ have the same shape.  Whenever we form a irrelevant function type $\forall x \of U.\, T[x]$, we require $T[x]$ to be shape-irrelevant in $x$.  This is the middle ground between ICC$^*$, where no restriction is placed on $T$ but $\eta$ for unit types is out of reach (at least for the moment), and Pfenning's \citeyearpar{pfenning:intextirr} irrelevance modality, adapted to full dependent types by \citet{abelScherer:types10}, which requires $T$ to be irrelevant in $x$ and, thus, has type equality $T[i] = T[j]$.

@@

The notation $\epiT U T$ is an
umbrella for three kinds of function types, where
$\evar \in \{ \erased, \noterased \}$ is a relevance annotation
borrowed from \citet{pfenning:intextirr}. $\erpiT[\noterased] U T$ is
the ordinary dependent function type, $\erpiT[\noterased] \Size T$ is
relevant size quantification, and $\erpiT \Size T$ is irrelevant size
quantification.  We omit the ``$\noterased$''-markers
from $\Pi$ by default (and also in contexts $\Gamma$)
and write $\forallT T$ for $\erpiT \Size T$.
%
Examples for relevant size quantification $\piT \Size T$ are
$\piT \Size {\Set[0]}$ and
$\piT \Size {\piT {(\Nat{\ind 0})} {\Set[0]}}$.
In a syntax with named variables and non-dependent function type
they could be written as $\Size \to \Set[0]$ and
$\funT z \Size \Nat z \to \Set[0]$, resp.
%
An instance of irrelevant
quantification $\forall T$ would be
$\forall.\, \piT {(\Nat{\ind 0})} {(\Nat{\ind 1})}$
which is
$\forall z.\, \Nat z \to \Nat z$
in a named syntax.
Herein, $\Nat z$ denotes the type of natural numbers below $z$. The
expression $\Size$ is a possible instance of $U$ in $\epiT U T$, or a
possible type of a variable in a typing context $\Gamma$, but not a
first-class type, \ie, we cannot construct our own functions on sizes.

>>>>

Sized types \cite{hughesParetoSabry:popl96} delegate the checking for structural descent to the type system by annotating data types with a size parameter.  The type checker can then ensure that in recursive calls the size goes down, certifying termination.  In the simplest setting \cite{gimenez:typeBased,abel:lmcs07}, the size is just an upper bound on the tree height of the data structure; however, more sophisticated size annotations have also been considered \cite{xi:terminationHOSC,blanqui:rta04}.  Most sized type systems are non-dependent \cite{amadio:guardcondition,abelPientka:jfp16,bartheGregoireRiba:lernet08,bartheGregoireRiba:csl08,blanquiRiba:lpar06,dalLagoGrellois:esop17}, yet the combination of sized and dependent types has been studied as well \cite{blanqui:csl05,bartheGregoirePastawski:lpar06,gregoireSacchini:lpar10,sacchini:lics13,sacchini:flops14}.  However, to the best of our knowledge, no study combines \emph{higher-rank size polymorphism} with full-fledged dependent types.\footnote{\citet{xi:terminationHOSC} has first-class size polymorphism, but only indexed types, no universes or large eliminations.}

@@

The central technical observation is that reflection and reification do not need the precise type $T$, they work the same for any \emph{shape} $S$ of $T$.
We managed, while not introducing a new syntax for shapes, to define a relation $T \shape S$ on type values stating that type $S$ qualifies as shape for type $T$.
Shapes unfold dynamically during reflection and reification.
For example, when reflecting a variable $x$ into the polymorphic function type $\forall i.\, F\,i$ where $F\,i = \Nat\,i \to \Nat\,i$, we obtain
$(\up{\forall i.\/ F\,i}x)\,i = \up{F\,i} (x\,i)$ for size $i$ and
$(\up{\forall i.\/ F\,i}x)\,j = \up{F\,j} (x\,j)$ for size $j$.  The new types $F\,i$ and $F\,j$ we reflect at are no longer equal (and they are not subtypes of each other), but they still have the same shape, $\Nat\_ \to \Nat\_$.  This means they will still move in lock-step in respect to $\eta$-expansion, which is sufficient to prove NbE correct for judgmental equality.
We call the enabling property of $F$ \emph{shape irrelevance}, meaning that for any pair $i$, $j$ of legal arguments, $F\,i$ and $F\,j$ have the same shape.  Whenever we form a irrelevant function type $\forall x \of U.\, T[x]$, we require $T[x]$ to be shape-irrelevant in $x$.  This is the middle ground between ICC$^*$, where no restriction is placed on $T$ but $\eta$ for unit types is out of reach (at least for the moment), and Pfenning's \citeyearpar{pfenning:lics01} irrelevance modality, adapted to full dependent types by \citet{abelScherer:types10}, which requires $T$ to be irrelevant in $x$ and, thus, has type equality $T[i] = T[j]$.

@@

The notation $\epiT U T$ is an
umbrella for three kinds of function types, where
$\evar \in \{ \erased, \noterased \}$ is a relevance annotation
borrowed from \citet{pfenning:lics01}. $\erpiT[\noterased] U T$ is
the ordinary dependent function type, $\erpiT[\noterased] \Size T$ is
relevant size quantification, and $\erpiT \Size T$ is irrelevant size
quantification.  We omit the ``$\noterased$''-markers
from $\Pi$ by default (and also in contexts $\Gamma$)
and write $\forallT T$ for $\erpiT \Size T$.
%
Examples for relevant size quantification $\piT \Size T$ are
$\piT \Size {\Set[0]}$ and
$\piT \Size {\piT {(\Nat{\ind 0})} {\Set[0]}}$.
In a syntax with named variables and non-dependent function type
they could be written as $\Size \to \Set[0]$ and
$\funT z \Size \Nat z \to \Set[0]$, resp.
%
An instance of irrelevant
quantification $\forall T$ would be
$\forall.\, \piT {(\Nat{\ind 0})} {(\Nat{\ind 1})}$
which is
$\forall z.\, \Nat z \to \Nat z$
in a named syntax.
Herein, $\Nat z$ denotes the type of natural numbers below $z$. The
expression $\Size$ is a possible instance of $U$ in $\epiT U T$, or a
possible type of a variable in a typing context $\Gamma$, but not a
first-class type, \ie, we cannot construct our own functions on sizes.
